Neural models for computer-assisted musical orchestration: a preliminary study

### 2020-05-14 paper/latex/paper.tex
Final revisions

The approach studied in \cite{Cella19} consists in finding a good orchestration for any given sound by successive iterations, exploring a feature space defined by a fixed feature function $f$. This method is currently considered state of the art and was implemented in Orch-Idea.\\

In this paper, we try a different approach, using Deep Neural networks. The main idea is to train a model to classify combinations of real instruments, and use it then for orchestration. By training it with real combinations of instruments, such as \{Flute C6, Bassoon C4, Bassoon G4\}, the model will hopefully acquire the ability to identify the presence of each instrument and its associaed pitch by building the appropriate filter. Thus, when a sound is given as input, those filters will identify what are the best instruments that match the target sound.\\

This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07}\cite{Kitahara05}\cite{Chetry05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16}\cite{Bian}\cite{Brunner}\cite{Defossez19}. \carmine{Can you check if those references are good to use in this context?}\\

@@

The features we use are the mel spectrogram of the input sound, because of its ability to represent the way we hear sounds and its common use in MIR\textbf{Reference}.\\

We found that even on a simplified version of this classification problem, parametric classifiers were unable to meaningfully identify instrument-note pairs as the number of instruments used in the combinations increased. Therefore the need for neural networks arose. The two architectures we used are a Convolutional Neural Network (CNN) with a Long Short-Term Memory (LSTM) unit, and ResNet, a well known residual architecture that already yielded good results for image classification \cite{He15}. We chose to use a CNN because of its success in audio classification \cite{Hershey16}. Our inclusion of an LSTM is because of its ability to learn long term dependencies in data \cite{Hochreiter97}, which is important given the temporal nature of audio.\\

@@

We used the TinySOL database to create our input data. TinySOL is a subset of the Studio On Line (SOL) database created by IRCAM \textbf{Reference needed}. TinySOL contains 1,529 samples from 12 instruments. The instruments come from different orchestral families: strings, woodwinds, and brass. Each sample is one instrument playing a single note in the ordinario style, with one of three dynamics: \textit{pp}, \textit{mf}, or \textit{ff}. The instruments and ranges over which they were recorded is summarized in Tab.\ref{tinysol}.  \carmine{Can you review this part about TinySOL and add more details if needed?}\\

@@

The choice of using the mel spectrogram as input features of the model is very common in music information \cite{Stevens37} and can be considered to be an accurate representation of the way humans perceive sound and music.\carmine{Do we need more references for mel spectrogram?}\\

@@

In order to increase variability in our generated data, we used two methods of data augmentation \cite{Salamon17}\cite{Bhardwaj17} : pitch shifting and partial feature dropout.\\

The first one was done on the TinySOL samples. We performed a slight pitch shift by shifting the frequency of the signal. We did this by reading the samples with different sample rates. Librosa automatically converts the data coming from the samples to match the specified sample rate \cite{mcfee15}. Thus, a small difference in sample rate will slightly modify the perceived pitch if played at the normal sample rate. In practice, the sample rates used for this data augmentation were within $5\%$ of the actual $44100$Hz.\\

@@

For orchestrating, we trained our models on combinations of 10 instruments: French Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet, and Clarinet. A target sound is input to the model, and the 10 classes with the highest probability are extracted. These 10 classes represent the instrument-note pairs that are most represented in the target, and can be from any combination of the 10 instruments.

@@

We evaluated our orchestrations both qualitatively and quantitatively. Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that represented a C\# minor chord, so we checked if the orchestration contained the notes of the chord. If a target had specific notes, we identified if the note or its partials were present. For example there is target of an Oboe playing an A4 and a Bassoon playing a C\#3, the solution from ResNet contained the partials of the Bassoon's note: C\#3, G\#4, C\#5, and E6 were all included on various instruments in the solution.

For quantitative evaluation, we used the distance metric defined in \eqref{distance} to calculate differences in timbre between targets and solutions. The equation takes in the full FFT of the target and full FFT of the solution. Then for each bin of the FFT, it calculates the absolute difference between the values. The differing values of $\lambda_1$ and $\lambda_2$ allow the metric to penalize the solution in different ways.

In the summations, all the terms with a $\lambda_1$ will be distances calculated when there was more energy in the target than the solution, since $\delta_{k1} = 1$ only when $x_k \ge \tilde{x}_k$. By the same reasoning, all the terms with a $\lambda_2$ are distances when the solution provided more energy to a frequency than the target; it overshot the frequency. Therefore, the relation between $\lambda_1$ and $\lambda_2$ determines whether a solution is penalized more for undershooting or overshooting the target.

@@

In order to have comparable solutions, we did not allow OrchIdea to create sparse solutions, meaning that OrchIdea was forced to use all 10 instruments in each solution. This creates a more fair comparison since our model is unable to create sparse solutions.

@@

Classification through deep learning seems promising for orchestration because of our model's ability to classify individual instruments and pitches out of dense combinations of samples. While our model is not able to outperform OrchIdea, it shows consistent results. When comparing our methods to OrchIdea, it is important to note that many of processes OrchIdea uses to optimize its solution are not present in our model. Our model also lacks the implementation of symbolic constraints, which is an important part of assisted orchestration.\\

@@

Future steps in this project include implementing various methods that are present in OrchIdea.\\

Our current model orchestrates all targets using the same number of samples, and this does not take into account the density of different targets. The solution to this is to allow sparse solutions in which the model decides how many samples should be used to best represent the target. This allows a small number of samples to be used for sonically sparse sounds and many to be used for sonically dense sounds.\\

>>>>

The approach studied in \cite{Cella19} consists in finding a good orchestration for any given sound by successive iterations, exploring a feature space defined by a fixed feature function $f$. This method is currently considered state of the art and was implemented in Orch-Idea.

In this paper, we try a different approach, using Deep Neural networks. The main idea is to train a model to classify combinations of real instruments, and use it then for orchestration. By training it with real combinations of instruments, such as \{Flute C6, Bassoon C4, Bassoon G4\}, the model will hopefully acquire the ability to identify the presence of each instrument and its associaed pitch by building the appropriate filter. Thus, when a sound is given as input, those filters will identify what are the best instruments that match the target sound.

This method is motivated by the good results obtained in previous research on musical instruments identification \cite{Benetos07}\cite{Kitahara05}\cite{Chetry05} and the more recent use of deep neural networks for musical classification \cite{lostanlen16}\cite{Bian}\cite{Brunner}\cite{Defossez19}. \carmine{Can you check if those references are good to use in this context?}

@@

The features we use are the mel spectrogram of the input sound, because of its ability to represent the way we hear sounds and its common use in MIR\textbf{Reference}.

We found that even on a simplified version of this classification problem, parametric classifiers were unable to meaningfully identify instrument-note pairs as the number of instruments used in the combinations increased. Therefore the need for neural networks arose. The two architectures we used are a Convolutional Neural Network (CNN) with a Long Short-Term Memory (LSTM) unit, and ResNet, a well known residual architecture that already yielded good results for image classification \cite{He15}. We chose to use a CNN because of its success in audio classification \cite{Hershey16}. Our inclusion of an LSTM is because of its ability to learn long term dependencies in data \cite{Hochreiter97}, which is important given the temporal nature of audio.

@@

We used the TinySOL database to create our input data. TinySOL is a subset of the Studio On Line (SOL) database created by IRCAM \textbf{Reference needed}. TinySOL contains 1,529 samples from 12 instruments. The instruments come from different orchestral families: strings, woodwinds, and brass. Each sample is one instrument playing a single note in the ordinario style, with one of three dynamics: \textit{pp}, \textit{mf}, or \textit{ff}. The instruments and ranges over which they were recorded is summarized in Tab.\ref{tinysol}.  \carmine{Can you review this part about TinySOL and add more details if needed?}

@@

The choice of using the mel spectrogram as input features of the model is very common in music information \cite{Stevens37} and can be considered to be an accurate representation of the way humans perceive sound and music.\carmine{Do we need more references for mel spectrogram?}

@@

In order to increase variability in our generated data, we used two methods of data augmentation \cite{Salamon17}\cite{Bhardwaj17} : pitch shifting and partial feature dropout.

The first one was done on the TinySOL samples. We performed a slight pitch shift by shifting the frequency of the signal. We did this by reading the samples with different sample rates. Librosa automatically converts the data coming from the samples to match the specified sample rate \cite{mcfee15}. Thus, a small difference in sample rate will slightly modify the perceived pitch if played at the normal sample rate. In practice, the sample rates used for this data augmentation were within $5\%$ of the actual $44100$Hz.

@@

For orchestrating, we trained our models on combinations of 10 instruments: French Horn, Oboe, Violin, Viola, Cello, Flute, Trombone, Bassoon, Trumpet, and Clarinet. This leads to 424 classes, each one representing an instrument-note pair. A target sound is input to the model, and the 10 classes with the highest probability are extracted. These 10 classes are the instrument-note pairs that are most represented in the target, and can be from any combination of the 10 instruments.

@@

We evaluated our orchestrations both qualitatively and quantitatively. We compared our solutions to OrchIdea, the state-of-the-art program for computer-assisted orchestration. In order to have comparable solutions, we did not allow OrchIdea to create sparse solutions, meaning that OrchIdea was forced to use all 10 instruments in each solution. This creates a more fair comparison since our model is unable to create sparse solutions.

Qualitative evaluation was done through an acoustic inspection of the solution, paying close attention to timbre and pitch. For targets that had harmonic content, it was noted if the partials present in the target were also represented in the orchestrated solution. For example, one of the samples of a bell had partials that represented a C\# minor chord, so we checked if the orchestration contained the notes of the chord. If a target had specific notes, we identified if the note or its partials were present. For example there is target of an Oboe playing an A4 and a Bassoon playing a C\#3, the solution from ResNet contained the partials of the Bassoon's note: C\#3, G\#4, C\#5, and E6 were all included on various instruments in the solution.

For quantitative evaluation, we used the distance metric defined in Eqn. \eqref{distance} to calculate differences in timbre between targets and solutions. The equation takes in the full FFT of the target and full FFT of the solution. Then for each bin of the FFT, it calculates the absolute difference between the values. The differing values of $\lambda_1$ and $\lambda_2$ allow the metric to penalize the solution in different ways.

In the first summation, $\lambda_1$ is multiplied by all the distances calculated when there was more energy in the target than the solution, since $\delta_{k1} = 1$ only when $x_k \ge \tilde{x}_k$. Similarly in the second summation, $\lambda_2$ is multiplied by all the distances when the solution provided more energy to a frequency than the target; it overshot the frequency. Therefore, the relation between $\lambda_1$ and $\lambda_2$ determines whether a solution is penalized more for undershooting or overshooting the target.

@@

@@

Classification through deep learning seems promising for orchestration because of our model's ability to classify individual instruments and pitches out of dense combinations of samples. While our model is not able to outperform OrchIdea, it shows consistent results. When comparing our methods to OrchIdea, it is important to note that many of processes OrchIdea uses to optimize its solution are not present in our model. Our model also lacks the implementation of symbolic constraints, which is an important part of assisted orchestration.

@@

Future steps in this project include implementing various methods that are present in OrchIdea.

Our current model orchestrates all targets using the same number of samples, and this does not take into account the density of different targets. The solution to this is to allow sparse solutions in which the model decides how many samples should be used to best represent the target. This allows a small number of samples to be used for sonically sparse sounds and many to be used for sonically dense sounds.
